---
title: "Ankle draft risk model under winter clothing conditions"
authors:
  - name: Tobias Kramer
    affiliations:
      - ref: cbe
    corresponding: true
    email: t.kramer@berkeley.edu
    orcid: 0000-0000-0000-0000
  - name: Junmeng Lyu
    affiliations:
      - ref: cbe
      - ref: sjtu
    orcid: 0000-0000-0000-0000
  - name: Stefano Schiavon
    affiliations:
      - ref: cbe
      - ref: arch
      - ref: eng
    orcid: 0000-0003-1285-5682 

affiliations:
  - id: cbe
    name: Center for the Built Environment, University of California, Berkeley, USA
  - id: arch
    name: Department of Architecture, College of Environmental Design, University of California, Berkeley, USA
  - id: eng
    name: Department of Civil and Environmental Engineering, College of Engineering, University of California, Berkeley, USA
  - id: sjtu
    name: Department of Architecture, School of Design, Shanghai Jiao Tong University, Shanghai, China

filters:
  - authors-block

date: "`r Sys.Date()`"
format:
  docx:
    link-citations: true
    reference-doc: template.docx
    toc: false
    number-sections: true
  html:
    toc: true
    number-sections: true
    theme: cosmo
    comments:
      hypothesis: true
bibliography: refs.bib
csl: apa.csl
execute:
  echo: false
  warning: false
  error: false
---

```{r load-main-analysis}
#| message: false
#| warning: false
#| paged-print: false
source(here::here("src", "R", "x_analysis.R"))
```

# Abstract {.unnumbered}

The ankle draft model incorporated in ASHRAE Standard 55 was derived
from studies in which participants mainly wore summer clothing with
exposed ankles (\~0.5 clo), potentially leading to overly conservative
perimeter heating or facade insulation requirements in winter conditions
where occupants typically wear long trousers, socks, and closed-toe
shoes (\~0.75 clo). This study evaluated thermal comfort responses to
ankle-level draft under such winter clothing conditions through
controlled laboratory experiments with 51 participants. Using a
within-subjects repeated-measures design, we exposed participants to
nine conditions varying supply air temperature (15, 17, and 19°C) and
ankle-level air speed (0.1–0.7 m/s) simulating window downdraft while
they wore clothing ensembles targeting 0.75 clo.

The existing ankle draft model systematically overpredicted
dissatisfaction across all conditions. While prior studies with exposed
ankles reported dissatisfaction rates of 23–57%, we observed rates of
only 6–20%, with the existing model overestimating dissatisfaction by an
average of 2.8 times and by as much as 39% at high air speeds. We fit
generalized linear mixed-effects models to data from this study and
prior research, developing separate prediction equations for
ankle-covered and ankle-uncovered conditions. The covered-ankle model
exhibited both a lower baseline dissatisfaction risk and reduced
sensitivity to air speed. Whole-body thermal sensation and ankle-level
air speed were the only significant predictors; ankle air temperature
and room temperature were not significant within the tested ranges.

These findings support differentiating ASHRAE Standard 55 ankle draft
criteria by clothing insulation level. Adopting a covered-ankle
criterion for winter heating scenarios will enable meaningful reductions
in perimeter heating and facade insulation requirements without
compromising occupant comfort, with associated benefits for installation
costs, energy use, and carbon emissions.

# Introduction

Cold downdraft near windows is a common source of local thermal
discomfort in buildings during winter [@heiselbergDraughtRiskCold1994;
@fangerAirTurbulenceSensation1988]. When outdoor temperatures drop, the
interior surface of windows, particularly large or poorly insulated
glazing, becomes significantly cooler than the surrounding air. This
temperature differential causes nearby air to cool, increase in density,
and descend, creating a localized airflow pattern that can cause
discomfort at the lower extremities of seated occupants
[@heiselbergDraughtRiskCold1994;
@larssonExperimentalInvestigationDowndraught2002]. Building designers
typically address this issue by increasing the insulation level of
windows, so that the interior surface temperature is higher, adding
mullions to break the downward flow and by installing perimeter heating
systems, such as radiators or in-floor heating, to counteract the
downdraft effect and to prevent thermal discomfort in the ankle region.

The current approach to evaluating ankle draft risk relies on a
predictive model presented in @liuPredictedPercentageDissatisfied2017,
which estimates the predicted percentage dissatisfied (PPD) with ankle
draft as a function of air speed and temperature at ankle level, as well
as whole-body thermal sensation. This model was derived from laboratory
experiments conducted under cooling conditions typical of underfloor air
distribution (UFAD) and displacement ventilation systems and is
incorporated into ASHRAE Standard 55 [@ashraeANSIASHRAEStandard2023].
Importantly, the 110 participants in these foundational studies,
including an earlier investigation by
@schiavonSensationDraftUncovered2016, wore summer clothing with exposed
ankles, resulting in clothing insulation levels of approximately 0.5
clo.

However, building occupants in winter typically wear substantially
different attire. Analysis of the ASHRAE Global Thermal Comfort Database
II [@foldvarylicinaDevelopmentASHRAEGlobal2018] indicates that median
clothing insulation in commercial spaces during heating seasons is
approximately 0.75 clo, with occupants commonly wearing long trousers,
closed-toe shoes and socks that cover the ankle region. This discrepancy
raises an important question: does the existing ankle draft model,
calibrated for summer conditions where in about two thirds of test cases
subjects had exposed ankles, accurately predict discomfort when ankles
are covered by clothing?

If the current model overestimates draft risk under winter clothing
conditions, the practical implications are significant. Designers may be
specifying solutions, like perimeter heating systems, that are larger
than necessary, or installing them in situations where they could be
avoided entirely. This has consequences for both first costs and
operational energy consumption and related carbon emission. Conversely,
if the model underestimates risk, occupants may experience discomfort
that compromises satisfaction and performance.

The objective of this study is to evaluate thermal comfort responses to
ankle-level draft under typical winter clothing conditions through
controlled laboratory experiments. We aim to: (1) characterize
whole-body and local thermal perception across a range of supply air
temperatures and air speeds representative of window downdraft
conditions; (2) compare observed dissatisfaction rates with predictions
from the existing ankle draft model; and (3) assess whether the current
ASHRAE 55 criteria require adjustment for winter scenarios. The findings
are intended to inform guidance on perimeter heating requirements and
support more energy-efficient building design without compromising
occupant comfort.

# Methods {#sec-methods}

## Experimental design

We conducted a controlled laboratory experiment using a within-subjects
repeated-measures design. Each participant was exposed to nine
experimental conditions in a 3×3 factorial arrangement, varying supply
air temperature (SAT) at three levels (15, 17, and 19°C) and air speed
at ankle level at three levels (low, medium, high; ranging between
0.1–0.7 m/s). The SAT and air speeds were selected to represent the
range of conditions associated with window downdraft under typical
winter conditions [@manzAnalysisThermalComfort2004] and to be consistent
with earlier studies focusing on summer conditions
[@liuPredictedPercentageDissatisfied2017;
@schiavonSensationDraftUncovered2016]

Participants attended four sessions: an introductory session for
explaining the experiment and obtaining consent, followed by three
experimental sessions of approximately two hours each. Each experimental
session corresponded to one SAT level, with participants experiencing
all three air speed conditions within that session in a counterbalanced
order. We scheduled sessions on separate days to minimize carryover
effects and subjects were allowed to attend the three sessions in their
preferred order.

## Climate chamber and apparatus

We conducted the experiment in the climate-controlled chamber at the
Center for the Built Environment, UC Berkeley. The chamber floor area is
approximately 25 m². We set up three workstations to allow simultaneous
testing of multiple participants and three seats for reacclimatization
during breaks in the back of the room. Ambient room conditions were
maintained at approximately 22°C operative temperature, 50% relative
humidity, and background air velocity below 0.1 m/s, targeting a
whole-body thermal sensation of approximately −0.5 (slightly cool) on
the ASHRAE seven-point scale as calculated using the CBE Comfort Tool
[@tartariniCBEThermalComfort2020].

![Experimental configuration and temporal protocol: (a) Seated occupant
positioned 0.7 m from the custom displacement diffusor, which simulates
window downdraft by directing conditioned air toward the ankle region
from behind; (b) Timeline of the 120-minute experimental sessions
featuring overlapping participant groups and synchronized questionnaire
intervals to characterize the progression of local and whole-body
thermal perception.](figs/experimental_setup.png){#fig-setup
width="75%"}

We used custom built displacement diffusors to deliver conditioned air
to the ankle region of participants seated at the three workstations.
The diffuser was positioned 0.7 m behind each workstation, directing
airflow toward the participants' lower legs from behind and simulating
the approach direction of window downdraft. The diffuser was connected
to an independent air handling unit capable of delivering supply air at
the target temperatures and flow rates. The three supply air
temperatures represent temperatures at the VAV outlet and we controlled
air flow rates by damper adjustment. Turbulence intensity at the
measurement location was approximately 30% (see @tbl-env), consistent
with values reported in prior ankle draft studies
[@liuPredictedPercentageDissatisfied2017]. The three seats in the back
of the room were not affected by the diffusers targeting the three
workstations only.

## Monitoring

### Environmental

We measured head-level air temperature (1.1 m) at each workstation using
calibrated Atmocube monitors (Atmotech Inc., United States; air
temperature: -40 to +125°C, ±0.5°C, relative humidity: 0 to 100%RH, ±2
%RH, CO~2~: 0 to 5000 ppm, ±50 ppm + 2.5% of reading at standard room
levels) positioned within 0.5 m of participants. To measure air velocity
and air temperature at ankle level (0.1 m) we used the AirDistSys 5000
omnidirectional hot-wire anemometers (Sensor Electronic, Poland, air
velocity: 0.05-5 m/s, accuracy: ±0.02 m/s ±2% of reading; temperature:
-10 to +50°C, accuracy: ±0.2°C, sampling interval: 1 s). Room-level
measurements of CO~2~ concentration, PM~2.5~, illuminance and sound
pressure levels were recorded throughout each session but not reported
here because they stayed constant and within expected ranges.

### Physiological

In addition, we recorded continuously participants skin temperature on
the lateral lower leg (mid-calf) and feet using iButton sensors
(Thermochron, Maxim Integrated, USA, Type DS1923; measurement range:
20°C to 85°C; accuracy: ±0.5°C; resolution: 0.0625°C) attached with
medical-grade hypoallergenic tape. We used the skin temperature data to
assess local cooling responses and as physiological correlates of
subjective thermal sensation.

### Subjective feedback {#sec-subjective-feedback}

To assess the perception of the thermal environment we used validated
scales consistent with prior ankle draft research and ASHRAE Standard 55
requirements. We employed two questionnaires: a comprehensive one,
encompassing all questions, administered shortly before the conclusion
of the exposure (F), and a concise, partial one, comprising a subset of
the questions, at the end of each adaptation period. For details on the
timing of the questionnaires, please refer to @fig-setup, and for
information on the constructs and their respective scales, please refer
to @tbl-scales below. We administered the questions electronically via a
custom survey tool written in Java. Participants were not required to
respond to any individual question, though completion rates were
monitored.

```{r}
#| label: tbl-scales
#| tbl-cap: "Survey constructs and measurement scales, including information if the question was used in the full (F) and partial (P) subject feedback questionnaire."
#| tbl-colwidths: [15,10,25,15,35]
#| echo: false

# filtering for only the question we actually analyze
readr::read_csv("tables/survey_scales.csv") |>
  slice(1:4, 7:8, 10) |> 
  knitr::kable(align = "l")
```

## Participants

We recruited
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(n)`
participants from the university community and local area through the
Experimental Social Science Laboratory (Xlab) platform and email
outreach to individuals who had participated in previous studies and
consented to future contact. Eligible participants were between 21 and
55 years of age, fluent in English, and free from chronic
cardiorespiratory conditions or recent major surgery. Pregnant
individuals were excluded. The UC Berkeley Committee for Protection of
Human Subjects (CPHS) approved the research protocol (CPHS
##2025-07-18795). Before the tests, all subjects signed an informed
consent form. Demographic details of the cohort are summarized in
@tbl-demographics.

## Experimental protocol

Upon arrival for each experimental session, we checked participants for
elevated temperature and fitted them with skin temperature sensors.
Participants then entered the climate chamber and completed a 20-minute
adaptation period in the acclimatization zone under baseline conditions.

To simulate typical winter attire, we instructed participants to wear a
long-sleeve top, long trousers (ending just above the ankle), thin
socks, and closed-toe shoes. This ensemble targets a clothing insulation
of approximately 0.75 clo, consistent with the median value observed in
commercial buildings during heating seasons
[@foldvarylicinaDevelopmentASHRAEGlobal2018]. Participants were
permitted to make minor clothing adjustments during sessions (e.g.,
rolling sleeves), and any changes were recorded.

Following adaptation, participants were exposed to three consecutive
20-minute conditions at different air speed levels, with 20-minute rest
periods between conditions (see @fig-setup). The sequence of air speed
conditions was counterbalanced across participants and sessions. At the
end of each exposure and adaptation interval, participants completed the
standardized questionnaires listed in @sec-subjective-feedback.

## Statistical analysis

### Variable processing

Thermal sensation votes were collected on a continuous scale. For
descriptive analyses, we mapped these values to the corresponding
categorical sensation levels, whereas the original continuous values
were retained for statistical modeling. We dichotomized ankle airflow
acceptability as acceptable or unacceptable for subsequent analyses. The
original recorded votes for thermal acceptability, thermal preference,
ankle thermal preference, and ankle airflow preference were retained
without further processing. Consistent with the definition used by
@liuPredictedPercentageDissatisfied2017, we defined ankle draft
dissatisfaction as an ankle thermal sensation vote below 0 (neutral)
together with an ankle airflow acceptability vote below 0.

### Within-subject comparisons

We used a within-subject repeated-measures design and conducted
comparisons between experimental conditions within participants. We
analyzed continuous variables using paired-samples t-tests and reported
paired Cohen's d as the effect size. We analyzed ordinal variables using
the paired Wilcoxon signed-rank test, reporting the test statistic based
on the normal approximation and the effect size r. For multiple pairwise
comparisons within each family of tests, we adjusted p-values using the
Benjamini-Hochberg method [@benjaminiHighberg1995].

### Evaluation of the existing ankle draft model

Applying it to the experimental data, we calculated the predicted
percentage of dissatisfaction with ankle draft at each observation
point. Predictions were then averaged by experimental condition and
compared with the observed dissatisfaction percentages. We used
calibration curves to quantitatively assess the model's predictive
performance in this study's dataset [@vancalster2016] and report the
calibration intercept and slope, AUC, and Brier score. All analyses were
implemented using the R package CalibrationCurves
[@deCockCalibrationCurves2023].

### Development of the new ankle draft model {#sec-method-new-model}

We modeled ankle draft dissatisfaction using a generalized linear mixed
model with a binomial distribution and logit link, as the outcome was
binary and repeated observations were correlated within subjects. We
included the subject as a random intercept to account for
between-subject heterogeneity and within-subject dependence arising from
repeated measurements. To enable an integrated analysis of the effects
of different clothing conditions on ankle draft dissatisfaction, we also
included data from @liuPredictedPercentageDissatisfied2017 in the
modeling dataset. Continuous predictors were mean-centered using the
overall means of the combined dataset before model fitting.

In this generalized linear mixed model, the inverse-logit transformation
is nonlinear. As a result, direct back-transformation of the
fixed-effects linear predictor yields a conditional probability,
representing an individual-level prediction when the random effect is
set to zero, rather than a population-averaged probability. To
characterize the average risk of dissatisfaction at the population
level, we therefore use marginal probabilities instead of conditional
probabilities [@neuhausKalbfleischHauck1991]. Specifically, we extracted
fixed-effect estimates and the variance of the random intercept from the
fitted model, assuming that the random intercept followed a normal
distribution with mean zero. We then approximated integration over the
random effects using Monte Carlo simulation (K = 10,000, random seed =
1), and calculated the marginal predicted probability for each
combination of predictor values.

Because these marginal probabilities generally cannot be written as a
simple closed-form equation, we used the Monte Carlo-simulated marginal
probabilities to fit an approximate linear equation for logit(P).
Applying the inverse-logit transformation to the fitted equation yielded
a closed-form prediction equation. We evaluated the accuracy of this
approximation using coefficient of determination (R^2^).

## Software and data

We performed all data processing, statistical analyses, and model
fitting in R (version 4.4.1). The main packages we used were tidyverse
(version 2.0.0), lme4 (version 2.0.1), rstatix (version 0.7.3), and
CalibrationCurves (version 3.0.0). Data and analysis code are available
on: https://github.com/CenterForTheBuiltEnvironment/ankle-draft-II

# Results {#sec-results}

## Participants and environmental conditions

```{r}
#| label: tbl-demographics
#| tbl-cap: "Summary of demographic information (median [IQR]) on study participants."

knitr::kable(demographic_d,
             col.names = c('Sex', 'No. Subjects', 'Age [a]', 'Height [m]', 'Weight [kg]', 'Body Mass Index (BMI)')
)

```

@tbl-demographics summarizes the demographic information of the study
subjects. A total of
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(n)`
subjects participated in the study
(`r demographic_d %>% dplyr::filter(sex == "Female") %>% dplyr::pull(n)`
female,
`r demographic_d %>% dplyr::filter(sex == "Male") %>% dplyr::pull(n)`
male,
`r demographic_d %>% dplyr::filter(sex == "Third Gender / Other") %>% dplyr::pull(n)`
other), with a median age of
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(age)`
years.

```{r}
#| label: tbl-env
#| tbl-cap: "Summary of environmental measurements."

knitr::kable(env_airflow_t,
             col.names = c('Measurement', '15°C', '17°C', '19°C')
)
```

@tbl-env provides an overview of the measured environmental conditions.
Across all three sessions, head-level air temperature remained stable at
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("15C")`
°C,
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("17C")`
°C, and
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("19C")`
°C, and relative humidity was comparable across sessions, confirming
stable background thermal conditions and establishing baseline in line
with the PMV target level of \~ -0.5 (slightly cool).

As the primary experimental manipulation, we set supply air temperatures
to 15°C, 17°C, and 19°C to simulate the effect of window downdraft.
Ankle-level temperatures were slightly higher than the supply
temperature due to mixing with the warmer ambient air, but remained
stable across all sessions. Within each session, we exposed subjects to
three airspeed levels at separate workstations. We measured consistent
air speeds across all three workstations and sessions, confirming
reliable reproduction of the low, medium, and high airspeed conditions.
Turbulence intensities remained low to medium throughout, indicating
stable and controlled airflow [@fangerAirTurbulenceSensation1988].

## Whole body thermal perception

```{r}
#| label: whole-body-helpers
#| echo: false

# Helper functions for inline text
get_sensation_pct <- function(sat, ws, response) {
  thermal_sensation_overall_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

get_preference_pct <- function(sat, ws, response) {
  thermal_preference_overall_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

# Aggregated summaries for text
sensation_neutral_cool <- thermal_sensation_overall_summary %>%
  dplyr::filter(response_value %in% c("Neutral", "Slightly cool", "Cool")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

# for ankle level comparison
sensation_slightlycool_cold <- thermal_sensation_overall_summary %>%
  dplyr::filter(response_value %in% c("Slightly cool", "Cool", "Cold")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

preference_no_change <- thermal_preference_overall_summary %>%
  dplyr::filter(response_value == "No change")

preference_warmer <- thermal_preference_overall_summary %>%
  dplyr::filter(response_value == "Warmer")
```

@fig-thermal-sensation-overall shows whole-body thermal sensation votes
plotted against ankle air temperature and air speed. Votes ranged
predominantly from Cool to Slightly warm across conditions. Thermal
sensation increased with supply air temperature in all air speed groups,
with steeper slopes at medium and high air speeds (b ≈
`r round(min(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c", group %in% c("medium", "high")) |> dplyr::pull(slope)), 2)`
–
`r round(max(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c", group %in% c("medium", "high")) |> dplyr::pull(slope)), 2)`
per °C) than at low air speed (b ≈
`r round(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c", group == "low") |> dplyr::pull(slope), 2)`
per °C); however, supply temperature explained limited variance across
all groups (R² ≤
`r round(max(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c") |> dplyr::pull(r2)), 2)`).
No systematic relationship was observed between whole-body thermal
sensation and ankle air speed, consistent with ankle-level airflow
having limited influence on whole-body thermal perception (see
Supplementary Information for response distributions).

![Whole-body thermal sensation votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear regression
lines and 95% confidence bands. Sensations were predominantly in the
Cool to Slightly warm range; supply temperature showed a modest positive
association at medium and high air speeds, while no clear relationship
with air speed was
observed.](figs/thermal_sensation_overall.png){#fig-thermal-sensation-overall
width="100%"}

```{r}
#| label: acceptability-helpers
#| echo: false

# Thermal acceptability summary
thermal_acceptability_summary <- thermal_acceptability_d %>%
  dplyr::mutate(acceptable = response_value > 0) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(
    n = dplyr::n(),
    pct_acceptable = mean(acceptable),
    mean_value = mean(response_value),
    .groups = "drop"
  )

acceptability_range <- thermal_acceptability_summary %>%
  dplyr::summarise(
    min = scales::percent(min(pct_acceptable), accuracy = 1),
    max = scales::percent(max(pct_acceptable), accuracy = 1)
  )

acceptability_by_sat <- thermal_acceptability_summary %>%
  dplyr::group_by(session_sat) %>%
  dplyr::summarise(
    min = scales::percent(min(pct_acceptable), accuracy = 1),
    max = scales::percent(max(pct_acceptable), accuracy = 1),
    .groups = "drop"
  )
```

@fig-thermal-acceptability-overall shows whole-body thermal
acceptability plotted against ankle air temperature and air speed.
Responses were predominantly on the acceptable side of the scale across
all conditions. No strong linear relationship was found between thermal
acceptability and either supply temperature or air speed (R² ≤
`r round(max(lm_stats_d |> dplyr::filter(outcome == "thermal_acceptability_overall") |> dplyr::pull(r2)), 3)`
in all groups), indicating that the range of conditions tested did not
substantially compromise whole-body thermal comfort (see Supplementary
Information for response distributions).

![Whole-body thermal acceptability votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear regression
lines and 95% confidence bands. Acceptability was predominantly positive
across all conditions, with no strong linear association with supply
temperature or air
speed.](figs/thermal_acceptability_overall.png){#fig-thermal-acceptability-overall
width="100%"}

## Ankle level observations

### Thermal perception

```{r}
#| label: ankle-helpers
#| echo: false

# Helper functions for ankle-level inline text
get_ankle_sensation_pct <- function(sat, ws, response) {
  thermal_sensation_ankle_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

get_ankle_preference_pct <- function(sat, ws, response) {
  thermal_preference_ankle_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

# Aggregated ankle sensation on cool side (cold + cool + slightly cool)
ankle_sensation_cool_side <- thermal_sensation_ankle_summary %>%
  dplyr::filter(response_value %in% c("Cold", "Cool", "Slightly cool")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

# Ankle preference for warmer
ankle_preference_warmer <- thermal_preference_ankle_summary %>%
  dplyr::filter(response_value == "Warmer")
```

@fig-thermal-sensation-ankle presents ankle-level thermal sensation
votes plotted against ankle air temperature and air speed. Sensations at
the ankles were generally cooler than whole-body ratings. Supply
temperature showed a positive association with ankle sensation across
all air speed groups (b ≈
`r round(min(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "t_supply_c") |> dplyr::pull(slope)), 2)`
–
`r round(max(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "t_supply_c") |> dplyr::pull(slope)), 2)`
per °C), consistent with whole-body results. The relationship with ankle
air speed was negative across all supply temperatures and strongest at
15°C (b =
`r round(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group == "15C") |> dplyr::pull(slope), 2)`,
R² =
`r round(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group == "15C") |> dplyr::pull(r2), 2)`),
indicating that higher air speeds increased local cool sensations most
noticeably at the coldest supply temperature. The association was weaker
at 17°C and 19°C (b ≈
`r round(max(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group %in% c("17C", "19C")) |> dplyr::pull(slope)), 2)`
to
`r round(min(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group %in% c("17C", "19C")) |> dplyr::pull(slope)), 2)`,
R² ≤
`r round(max(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group %in% c("17C", "19C")) |> dplyr::pull(r2)), 2)`)
(see Supplementary Information for response distributions).

![Ankle-level thermal sensation votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear regression
lines and 95% confidence bands. Ankle sensations were generally cooler
than whole-body ratings; the strongest negative association with air
speed was observed at the 15°C supply
temperature.](figs/thermal_sensation_ankle.png){#fig-thermal-sensation-ankle
width="100%"}

```{r}
#| label: air-movement-helpers
#| echo: false

# Air movement acceptability summary
air_acceptability_acceptable <- air_movement_acceptability_summary %>%
  dplyr::filter(acceptability == "Acceptable")

air_acceptability_by_sat <- air_acceptability_acceptable %>%
  dplyr::group_by(session_sat) %>%
  dplyr::summarise(
    min = scales::percent(min(pct), accuracy = 1),
    max = scales::percent(max(pct), accuracy = 1),
    .groups = "drop"
  )

# Air movement preference for less
air_pref_less <- air_movement_preference_summary %>%
  dplyr::filter(acceptability == "Less")

air_pref_no_change <- air_movement_preference_summary %>%
  dplyr::filter(acceptability == "No change")
```

### Air movement perception

@fig-air-move-acceptability shows air movement acceptability at the
ankle level plotted against supply temperature and air speed. Despite
the localized cooling, responses were predominantly on the acceptable
side of the scale. Air movement acceptability showed a consistent
negative association with air speed across all supply temperatures (b ≈
`r round(max(lm_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "v_air_m_s") |> dplyr::pull(slope)), 2)`
to
`r round(min(lm_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "v_air_m_s") |> dplyr::pull(slope)), 2)`
per m/s), while the relationship with supply temperature was negligible
for the low and high air speed groups (b ≈
`r round(mean(lm_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "t_supply_c", group %in% c("low", "high")) |> dplyr::pull(slope)), 2)`)
and modest for the medium group (b ≈
`r round(lm_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "t_supply_c", group == "medium") |> dplyr::pull(slope), 2)`).
Linear regression explained a small proportion of variance in all groups
(R² ≤
`r round(max(lm_stats_d |> dplyr::filter(outcome == "air_movement_acceptability") |> dplyr::pull(r2)), 3)`),
indicating substantial individual variability in acceptability ratings
(see Supplementary Information for response distributions).

![Ankle-level air movement acceptability votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear regression
lines and 95% confidence bands. Acceptability was predominantly
positive; a consistent negative association with air speed was observed
across all supply
temperatures.](figs/air_movement_acceptability_ankles.png){#fig-air-move-acceptability
width="100%"}

### Skin temperature

```{r}
#| label: tsk-delta-helpers
#| echo: false

tsk_delta_last1_range <- tsk_delta_last1 %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(mean_delta_last1_c = mean(mean_delta_last1_c, na.rm = TRUE), .groups = "drop") %>%
  dplyr::summarise(
    min_val = round(min(mean_delta_last1_c, na.rm = TRUE), 1),
    max_val = round(max(mean_delta_last1_c, na.rm = TRUE), 1)
  )

tsk_delta_sig_low_med_all <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "medium") %>%
  dplyr::summarise(all_sig = all(p.adj < 0.05, na.rm = TRUE)) %>%
  dplyr::pull(all_sig)

tsk_delta_sig_low_med_max_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "medium") %>%
  dplyr::summarise(max_p = max(p.adj, na.rm = TRUE)) %>%
  dplyr::pull(max_p)

tsk_delta_sig_low_high_all <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "high") %>%
  dplyr::summarise(all_sig = all(p.adj < 0.05, na.rm = TRUE)) %>%
  dplyr::pull(all_sig)

tsk_delta_sig_low_high_max_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "high") %>%
  dplyr::summarise(max_p = max(p.adj, na.rm = TRUE)) %>%
  dplyr::pull(max_p)

stopifnot(isTRUE(tsk_delta_sig_low_med_all))
stopifnot(isTRUE(tsk_delta_sig_low_high_all))

tsk_delta_sig_med_high <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "medium", group2 == "high") %>%
  dplyr::mutate(sig = p.adj < 0.05) %>%
  dplyr::filter(sig) %>%
  dplyr::pull(session_sat)

tsk_delta_sig_med_high_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "medium", group2 == "high", p.adj < 0.05) %>%
  dplyr::pull(p.adj)

format_p_threshold <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "p_adj < 0.001",
    p < 0.01 ~ "p_adj < 0.01",
    p < 0.05 ~ "p_adj < 0.05",
    TRUE ~ sprintf("p_adj = %.3f", p)
  )
}
```

@fig-tsk illustrates the progression of ankle skin temperatures
throughout the exposure period. Using the pre-exposure adaptation period
as a baseline, ankle skin temperatures dropped under all conditions,
with final reductions ranging from `r tsk_delta_last1_range$min_val`°C
to `r tsk_delta_last1_range$max_val`°C. In general, the magnitude of
this skin temperature reduction increased with air speed, as medium and
high air speeds produced greater local cooling than low air speed across
all three supply air temperatures.

Statistical analysis using BH correction confirmed that the cooling
differences between low and medium air speeds, as well as between low
and high air speeds, were significant at all supply temperatures
(`r format_p_threshold(tsk_delta_sig_low_med_max_p)` and
`r format_p_threshold(tsk_delta_sig_low_high_max_p)`, respectively). In
contrast, the difference between medium and high air speeds reached
significance only at `r tsk_delta_sig_med_high`
(`r sprintf("p_adj = %.3f", tsk_delta_sig_med_high_p)`). These findings
indicate that increasing local airflow increases the short-term cooling
response of the ankle skin and most noticeable temperature drops occur
when shifting away from the lowest air speed setting.

![Time course of ankle skin temperature during the formal exposure phase
across the nine experimental conditions for (a) absolute skin
temperature and (b) change relative to the adaptation-period baseline (
Shaded bands denote the between-subject mean ± 1 SD): Local cooling
intensified as air speed increased, with the most significant
temperature drops occurring when moving from low to medium
airflow.](figs/tsk_p.png){#fig-tsk width="90%"}

## Dissatisfaction with ankle draft {#sec-model-comp}

```{r}
#| label: setup-results-data
#| echo: false
#| message: false

# Converting decimals to percentages and rounding for cleaner inline text
res <- model_comparison_d %>%
  mutate(across(c(ppd_liu_mean, dissatisfied_rate, diff), ~ round(.x * 100, 1)))

liu_model_summary <- model_comparison_d %>%
  dplyr::summarise(
    mean_observed = round(mean(dissatisfied_rate) * 100, 0),
    mean_predicted = round(mean(ppd_liu_mean) * 100, 0),
    overprediction_ratio = round(mean(ppd_liu_mean) / mean(dissatisfied_rate), 1)
  )

liu_diff_by_speed <- model_comparison_d %>%
  dplyr::group_by(session_air_flow) %>%
  dplyr::summarise(mean_diff = mean(diff), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_diff))

liu_max_diff_speed <- liu_diff_by_speed %>% dplyr::slice(1) %>% dplyr::pull(session_air_flow)
liu_min_diff_speed <- liu_diff_by_speed %>% dplyr::slice(n()) %>% dplyr::pull(session_air_flow)
```

@fig-model-comp(a) compares the dissatisfaction rates predicted by the
Liu model against our experimental observations. Across all conditions,
the existing model consistently overpredicted local thermal
dissatisfaction. At a 15°C ankle temperature, the discrepancy between
predicted and observed rates widened as air speed increased: the model
overpredicted dissatisfaction by `r res$diff[1]` percentage points (pp),
rising to `r res$diff[3]` pp at high flow (0.6 m/s).

We observed similar trends at 17°C and 19°C. Notably, for the 19°C
condition at high air speed (0.6 m/s), the model predicted a PPD of
`r res$ppd_liu_mean[9]`%, while the actual dissatisfied rate was only
`r res$dissatisfied_rate[9]`%, resulting in a mean overestimation of
`r res$diff[9]` pp. Across all experimental sessions, the model
overpredicted dissatisfaction by an average of `r round(mean(res$diff))`
pp.

![Performance and calibration of the Liu local dissatisfaction model:
(a) The model consistently overpredicted ankle-level dissatisfaction
across all conditions, with discrepancies widening at higher air speeds
(points represent mean values, with the segment length indicating the
magnitude of model overestimation); (b) calibration curve of the
existing model on this study
dataset.](figs/model_error_combined.png){#fig-model-comp width="90%"}

In addition, we evaluated the performance of the existing ankle draft
model by fitting its predictions—based on local air speed and whole-body
thermal sensation—against our observed binary satisfaction outcomes
using logistic regression. The resulting calibration curve is shown in
@fig-model-comp(b). In a well-calibrated model, the curve should closely
follow the 45° ideal line; however, our results show the curve located
predominantly below this line.

While the model maintained acceptable discriminative ability (AUC =
0.77), meaning it can effectively rank which conditions are more likely
to cause discomfort, the calibration statistics reveal significant
inaccuracy in the magnitude of its predictions. With a Brier score of
0.14 and a systematic shift away from the ideal line, the results
confirm that the existing model consistently overestimates the
probability of ankle draft dissatisfaction under winter conditions.

## Updated model for winter conditions

```{r}
#| label: setup-updated-model
#| echo: false
#| message: false

model_input_coef <- summary(model_input)$coefficients

model_screening <- tibble::tibble(
  term = rownames(model_input_coef),
  estimate = model_input_coef[, "Estimate"],
  std_error = model_input_coef[, "Std. Error"],
  z_value = model_input_coef[, "z value"],
  p_value = model_input_coef[, "Pr(>|z|)"]
)

model_screening_table <- model_screening %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    term_label = dplyr::recode(
      term,
      "v_air_m_s" = "Air speed at ankle",
      "t_air_c" = "Room air temperature",
      "t_supply_c" = "Air temperature at ankle",
      "thermal_sensation" = "Whole-body thermal sensation",
      "turbulence_intensity" = "Turbulence intensity",
      "clothing_typeshort" = "Ankle condition (uncovered vs. covered)",
      "sourcetoby" = "Season"
    ),
    estimate_fmt = sprintf("%.2f", estimate),
    std_error_fmt = sprintf("%.2f", std_error),
    z_value_fmt = sprintf("%.2f", z_value),
    p_value_fmt = dplyr::case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) %>%
  dplyr::select(term_label, estimate_fmt, std_error_fmt, z_value_fmt, p_value_fmt)

model_formula_inline <- model_formula %>%
  dplyr::mutate(
    intercept_fmt = sprintf("%.3f", Intercept),
    v_fmt = sprintf("%.3f", V_coef),
    r2_fmt = sprintf("%.4f", R2),
    formula_text_expanded = sprintf(
      r"(PPD_{AD} = \frac{\exp\left(%s + %sV %s %sTS\right)}{1 + \exp\left(%s + %sV %s %sTS\right)})",
      intercept_fmt,
      v_fmt,
      ifelse(TS_coef < 0, "-", "+"),
      sprintf("%.3f", abs(TS_coef)),
      intercept_fmt,
      v_fmt,
      ifelse(TS_coef < 0, "-", "+"),
      sprintf("%.3f", abs(TS_coef))
    )
  )

model_exposed <- model_formula_inline %>% dplyr::filter(Model == "exposed")
model_covered <- model_formula_inline %>% dplyr::filter(Model == "unexposed")

glmm_exposed_coef <- lme4::fixef(m_glmm_exposed)
glmm_covered_coef <- lme4::fixef(m_glmm_unexposed)
glmm_delta_intercept <- glmm_covered_coef["(Intercept)"] - glmm_exposed_coef["(Intercept)"]

extract_final_model_table <- function(mod, model_label) {
  coef_mat <- summary(mod)$coefficients
  tibble::tibble(
    model = model_label,
    term = rownames(coef_mat),
    estimate = coef_mat[, "Estimate"],
    std_error = coef_mat[, "Std. Error"],
    z_value = coef_mat[, "z value"],
    p_value = coef_mat[, "Pr(>|z|)"]
  )
}

model_final_table <- dplyr::bind_rows(
  extract_final_model_table(m_glmm_exposed, "Ankle uncovered"),
  extract_final_model_table(m_glmm_unexposed, "Ankle covered")
) %>%
  dplyr::mutate(
    term_label = dplyr::recode(
      term,
      "(Intercept)" = "Intercept",
      "v_air_m_s" = "Air speed at ankle",
      "thermal_sensation" = "Whole-body thermal sensation"
    ),
    estimate_fmt = sprintf("%.2f", estimate),
    std_error_fmt = sprintf("%.2f", std_error),
    z_value_fmt = sprintf("%.2f", z_value),
    p_value_fmt = dplyr::case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) %>%
  dplyr::group_by(model) %>%
  dplyr::mutate(
    model_display = dplyr::if_else(dplyr::row_number() == 1, model, "")
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(model_display, term_label, estimate_fmt, std_error_fmt, z_value_fmt, p_value_fmt)
```

As shown in @sec-model-comp, the existing model showed limited accuracy
in predicting ankle draft dissatisfaction under winter conditions. This
highlights the need for a revised model that accounts for differences in
application contexts. To enable an integrated analysis of the effects of
different factors on ankle draft dissatisfaction, we combined data from
@liuPredictedPercentageDissatisfied2017 and collected in this study in
the modeling dataset. To identify significant predictors, a model
including all candidate predictors was first fitted following the
methods described in @sec-method-new-model. Given that both datasets
were obtained from repeated measurements, subject-level dependence was
accounted for in the analysis. The result of the screening model is
shown in @tbl-updated-model-screening. Ankle-level air speed, whole-body
thermal sensation, and ankle condition were significantly correlated
with ankle draft dissatisfaction, whereas room air temperature,
ankle-level air temperature, and turbulence intensity were not
significant within the range investigated in this study.

```{r}
#| label: tbl-updated-model-screening
#| tbl-cap: "Results of the generalized linear mixed-effects model used for predictor screening."
#| echo: false

knitr::kable(
  model_screening_table,
  col.names = c("Predictor", "Estimate", "SE", "z", "p")
)
```

```{r}
#| label: tbl-updated-model-final
#| tbl-cap: "Generalized linear mixed-effects models for the ankle-uncovered and ankle-covered conditions."
#| echo: false

knitr::kable(
  model_final_table,
  col.names = c("Condition", "Predictor", "Estimate", "SE", "z", "p")
)
```

Based on these results, separate mixed-effects models for ankle draft
dissatisfaction were developed for the ankle-uncovered and ankle-covered
conditions, using ankle-level air speed and whole-body thermal sensation
as predictors. As shown in @tbl-updated-model-final, the two final
models exhibited the same structure but different coefficients. The
intercept of the covered model was 1.91 units lower than that of the
uncovered model, indicating a lower baseline risk of ankle draft
dissatisfaction when the ankle was covered by clothing. In addition, the
air speed coefficient was smaller in the covered model than in the
uncovered model, indicating reduced sensitivity of dissatisfaction to
air speed under ankle-covered conditions.

For the ankle-uncovered condition, the approximate closed-form equation
derived from the marginal predictions of the mixed-effects model can be
written as:

$$
`r model_exposed$formula_text_expanded`
$$

For the ankle-covered condition, the corresponding approximate
closed-form equation is:

$$
`r model_covered$formula_text_expanded`
$$

Where, $PPD_{AD}$ is the predicted percentage dissatisfied with ankle
draft, $V$ is air speed at ankle (m/s), and $TS$ is whole-body thermal
sensation.

These linear approximations reproduced the marginal predictions of the
mixed-effects models with high accuracy. The corresponding $R^2$ values
were `r model_exposed$r2_fmt` for the uncovered condition and
`r model_covered$r2_fmt` for the covered condition, indicating that the
approximate equations preserved the predictive behavior of the original
mixed-effects models well. The resulting model predictions are shown in
@fig-updated-model.

![Updated ankle draft dissatisfaction model under exposed and covered
conditions. (a) ankle-uncovered conditions, and (b) ankle-covered
conditions.](figs/Model.png){#fig-updated-model width="60%"}

# Discussion

## Comparison with prior research

The central finding of this study is that ankle-level dissatisfaction
under winter clothing conditions (≥0.75 clo) is substantially lower than
predicted by the existing model of
@liuPredictedPercentageDissatisfied2017 and than observed in prior
studies with exposed ankles. In the experiments of
@schiavonSensationDraftUncovered2016, where participants wore summer
clothing (\~0.6 clo) with exposed ankles and lower legs, unacceptability
rates ranged from 23% to 57%. Under the ankle-covered conditions in this
study, airflow unacceptability ranged from only 6% to 20% across
comparable air speed and temperature conditions.

These findings align with other research on the protective effects of
lower-body coverage. @zhouDraughtSensationAssessment2022 reported that
high-top shoes increased ankle airflow acceptability by approximately
15% in stratified heating environments, demonstrating that even partial
ankle coverage meaningfully reduces sensitivity to local airflow.
Similarly, @toftumDraughtSensitivityInfluenced1996 found that occupants
with cooler overall thermal sensations are more likely to perceive
airflow as uncomfortable, and that higher clothing insulation attenuates
the interaction between ambient temperature and airflow-induced
discomfort.

The limited variance explained by the linear regression models in
@fig-thermal-sensation-ankle and @fig-air-move-acceptability further
supports this interpretation. Across all conditions, R² values did not
exceed `r round(max(lm_stats_d$r2), 2)`, with the strongest association
observed for ankle-level thermal sensation versus air speed at 15°C (R²
=
`r round(lm_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group == "15C") |> dplyr::pull(r2), 2)`).
@liuPredictedPercentageDissatisfied2017 reported an R² of 0.10 for ankle
air movement acceptability versus air speed alone, and an adjusted R² of
0.22 for a model including both whole-body thermal sensation and ankle
air speed — substantially higher than our values and consistent with the
greater sensitivity to ankle draft expected under summer clothing
conditions with exposed ankles.

Regarding predictor selection, only whole-body thermal sensation and air
speed were significant predictors of dissatisfaction in our models,
consistent with the structures reported by
@toftumDraughtSensitivityInfluenced1996 and
@liuPredictedPercentageDissatisfied2017.
@chengExperimentalStudyDraftrelated2023 similarly identified a
relationship between overall thermal sensation and draft rate
thresholds, further supporting its inclusion as a predictor.

Although the dataset used by @liuPredictedPercentageDissatisfied2017
included both exposed and covered ankle conditions, no significant
clothing effect was identified in their analysis even if the ankle
condition (covered vs uncovered) was the third most important variable
after ankle air speed and whole-body thermal sensation. The discrepancy
in the significance of this variable likely stems from their use of
standard logistic regression, which pooled observations across subjects
without accounting for the non-independence introduced by repeated
measures. This approach may have masked the clothing effect as a
between-subject variable due to inter-individual variability, which is a
limitation addressed in our study through the use of mixed-effects
models.

## Mechanisms

Two complementary pathways, physiological and psychological, may explain
the reduced sensitivity to ankle draft under winter clothing conditions.

From a physiological perspective, the skin temperature results indicate
that winter clothing attenuates, though does not eliminate, the local
cooling effect of airflow. During the final steady-state period,
significant reductions in ankle skin temperature occurred only at the
coldest supply temperature (15°C), where medium and high air speeds
produced lower skin temperatures than the low-speed condition. At 17°C
and 19°C, variations in air speed did not yield significant differences
in steady-state skin temperature, and no significant differences were
observed between medium and high air speeds at any supply temperature.
These findings suggest that detectable local cooling requires a
sufficiently large environmental contrast.

Thermal sensation changes are more closely associated with rapid
variations in skin temperature than with absolute values
[@ChoiInvestigationHumanBody2012; @KenshaloCoolThresholds1968]. In this
study, ankle $\Delta T_{\text{sk}}$ relative to the baseline adaptation
period showed significant differences between low and medium/high air
speed conditions across all supply temperatures (effect size
$d \approx 0.6\text{-}0.9$). Consistent with this, local ankle thermal
sensation at medium and high speeds was significantly lower than at low
speed under the 15°C condition. However, air movement acceptability did
not differ significantly across conditions, indicating that while air
speed produced measurable shifts in skin temperature and local thermal
sensation, the magnitude of these changes was insufficient to generate
distinguishable differences in dissatisfaction within the tested range.
From a heat transfer perspective, clothing coverage increases local
thermal resistance and reduces convective and radiative heat exchange
[@oguroEvaluationEffectAir2001; @nishiDirectMeasurementClothing1975],
suppressing both the magnitude and variability of skin temperature
changes and making it less likely that occupants cross discomfort
thresholds.

Psychological factors may also contribute to the observed tolerance.
Occupants wearing thicker lower-body clothing may perceive themselves as
"protected," increasing their tolerance to local airflow. Prior research
has demonstrated that thermal perception is not solely determined by
physical stimuli but is modulated by expectations and informational
cues. Providing occupants with temperature information or group
perception feedback can improve thermal comfort evaluations
[@zhangGroupPerceptionFeedback2022], and thermal expectations themselves
influence perception [@schweikerThermalExpectation2020]. Field studies
have also identified associations between perceived control and thermal
satisfaction in winter conditions [@xuPerceivedControlWinter2021]. These
findings suggest that the subjective sense of being protected by
clothing may enhance tolerance to ankle draft, though this psychological
pathway was not directly measured in the present study and warrants
further investigation.

Together, these physiological and psychological mechanisms likely act in
concert: clothing reduces the physical cooling stimulus while
simultaneously altering occupants' expectations and perceived
vulnerability to draft.

## Implications for building design and standards

The ankle draft model of @liuPredictedPercentageDissatisfied2017, now
incorporated into ASHRAE Standard 55, does not distinguish between
covered and uncovered ankle conditions. Given the standard's acceptable
thermal sensation range of PMV = −0.5 to +0.5, the current criterion
requires ankle-level air speed below 0.22 m/s to achieve less than 20%
dissatisfied when TS = −0.5. This constraint may be overly conservative
for typical winter environments where occupants wear clothing that
covers the ankle.

The findings of this study support a differentiated approach. For winter
conditions with ankle coverage (≥0.75 clo), the covered-ankle model
developed here indicates that dissatisfaction remains within acceptable
limits at air speeds that would exceed the current threshold. Empirical
predictions of window downdraft suggest that maximum velocities at
approximately 1 m from the façade are around 0.25 m/s, with peak
velocities often falling within 0.1–0.2 m/s under milder outdoor
conditions or with smaller window areas
[@lyonsWindowPerformanceHuman1999]. Many of these scenarios would be
acceptable under the covered-ankle model but flagged as problematic
under the current single-threshold approach.

For practitioners, this suggests a two-criteria framework: apply the
existing model (or the uncovered-ankle model from this study) when
evaluating spaces where occupants are likely to have exposed ankles, and
apply the covered-ankle model for winter heating scenarios where long
trousers, socks, and closed-toe shoes are expected. Within the
near-neutral thermal range where ASHRAE Standard 55 applies, PMV
provides a reasonable proxy for whole-body thermal sensation during
design evaluation.

The practical benefits are significant. Perimeter heating systems,
increased façade insulation and other techniques used to reduce the
downdraft represent non-trivial costs and carbon emissions in both
installation and operation. If covered and uncovered conditions are
evaluated separately, designers may be able to downsize or remove
solutions used to manage the problem. This approach could yield
meaningful reductions in first costs, energy consumption, and associated
carbon emissions while maintaining occupant comfort.

## Limitations

Several limitations should be considered when interpreting these
findings.

First, this study was conducted in a controlled laboratory environment
with standardized clothing, which differs from real settings where
occupants exhibit considerable variability in attire even during winter
months. While we targeted 0.75 clo to represent median winter clothing,
actual insulation levels vary with individual preferences, building
temperatures, and cultural norms. The protective effect of ankle
coverage may differ with clothing types not tested here, such as
ankle-length skirts, cropped trousers, or low-cut shoes.

Second, the participant sample consisted primarily of
university-affiliated individuals within a limited age range. Thermal
sensitivity varies with age, and older occupants may respond differently
to ankle draft. The generalizability of these findings to broader
occupant populations warrants further investigation.

Third, the experimental protocol involved 20-minute exposure periods,
whereas real office occupants experience sustained exposure over hours.
Whether dissatisfaction accumulates or diminishes with prolonged
exposure was not assessed. Similarly, participants remained seated
throughout; standing or walking occupants may have different thermal
experiences at the ankle level.

Fourth, the experimental conditions covered supply air temperatures of
15–19°C and air speeds of 0.1–0.7 m/s. While these ranges were selected
to represent typical window downdraft conditions, more extreme scenarios
(colder supply temperatures or higher velocities) were not tested. The
model predictions should be applied with caution outside the tested
parameter space.

Fifth, airflow was delivered from behind the participants to simulate
window downdraft approaching from the façade. Other airflow directions
(lateral or frontal) may produce different perceptual responses,
particularly given potential differences in skin exposure and attention
to the stimulus source.

Finally, the experimental setup simulated the convective component of
window downdraft but did not include radiant asymmetry from a cold
window surface. In real buildings, occupants near poorly insulated
glazing experience both convective cooling from downdraft and radiant
heat loss to the cold surface. The combined effect of these two
mechanisms on ankle-level comfort was beyond the scope of this study.

# Conclusions

This study evaluated thermal comfort responses to ankle-level draft
under typical winter clothing conditions. Through controlled laboratory
experiments with participants wearing approximately 0.75 clo ensembles
that covered the ankle, we characterized local and whole-body thermal
perception across a range of supply air temperatures (15-19°C) and air
speeds (0.1–0.7m/s) simulating window downdraft.

The principal findings indicate that the existing ankle draft model
systematically overpredicts dissatisfaction under winter clothing
conditions. While prior studies with exposed ankles reported
dissatisfaction rates between 23% and 57% under similar air speeds, our
observations with covered ankles ranged from only 6% to 20%. This
suggests that ankle coverage substantially reduces sensitivity to local
draft, a conclusion supported by the covered-ankle model developed in
this study; the model demonstrates both a lower baseline risk and a
reduced response to air speed compared to uncovered-ankle models, with
an intercept difference of approximately 1.5 units on the logit scale.

Consistent with prior model structures, we found that whole-body thermal
sensation and air speed at the ankle are the primary predictors of
dissatisfaction, whereas ankle air temperature, room temperature, and
turbulence intensity were not significant within the tested ranges. This
reduced sensitivity under winter clothing appears to stem from a
combination of physiological mechanisms, specifically attenuated skin
cooling due to increased thermal resistance, and potentially
psychological factors, such as a perceived sense of protection provided
by the clothing coverage.

These findings support a differentiated approach to ankle draft
assessment in building standards. For winter heating scenarios where
occupants typically wear clothing that covers the ankle (≥0.75 clo), the
covered-ankle model provides more accurate predictions than the current
ASHRAE Standard 55 criterion. Adopting separate criteria for covered and
uncovered conditions could enable significant reductions in perimeter
heating requirements without compromising occupant comfort, yielding
benefits in installation costs, energy consumption, and carbon
emissions.

# CRediT authorship contribution statement {.unnumbered}

**Tobias Kramer**: Conceptualization, Data curation, Formal analysis,
Methodology, Investigation, Project administration, Software, Writing -
original draft, Writing - review & editing. **Junmeng Lyu**: Data
curation, Formal analysis, Methodology, Investigation, Software,
Writing - original draft, Writing - review & editing. **Stefano
Schiavon**: Conceptualization, Funding acquisition, Methodology, Project
administration, Supervision, Writing - review & editing.

# Declaration of competing interest {.unnumbered}

All authors declare no competing interest. The Center for the Built
Environment (CBE) at the University of California, Berkeley, with which
the authors are affiliated, is advised and funded in part by
approximately 50 partners that represent a diversity of organizations
from the building industry, including manufacturers, building owners,
facility managers, contractors, architects, engineers, government
agencies, and utilities.

# Data availability {.unnumbered}

The datasets generated and analyzed during the current study as well as
the full source code of the data analysis workflow are available in the
following GitHub repository:
<https://github.com/CenterForTheBuiltEnvironment/ankle-draft-II>

# Declaration of Generative AI and AI-assisted technologies in the writing process {.unnumbered}

During the preparation of this work, the authors utilized the chatbot
Claude to enhance the language and readability of the content.
Subsequently, they reviewed and edited the material as required, while
maintaining full accountability for the publication’s content.

# Acknowledgements {.unnumbered}

Schüco and the Center for the Built Environment at the University of
California, Berkeley supported this research.

# Appendix A: Response distributions {.unnumbered}

The following figures show the full response distributions for thermal
sensation, thermal preference, and air movement acceptability and
preference. These complement the scatter plots with linear regression
lines presented in @sec-results in the main text.

![Whole-body thermal sensation (a) and thermal preference (b) votes by
supply air temperature and air speed. Participants predominantly
reported neutral to cool sensations, with cooler supply temperatures
increasing the proportion preferring warmer
conditions.](figs/supp_thermal_perception_overall.png){#fig-supp-thermal-perception-overall
width="75%"}

![Ankle-level thermal sensation (a) and thermal preference (b) votes by
supply air temperature and air speed. Sensations were generally cooler
at the ankles than whole-body ratings, with cooler supply temperatures
and higher air speeds increasing cool sensations and the preference for
warmer
conditions.](figs/supp_thermal_perception_ankle.png){#fig-supp-thermal-perception-ankle
width="75%"}

![Ankle-level air movement acceptability (a) and preference (b) votes by
supply air temperature and air speed. Most participants found ankle air
movement acceptable, though higher air speeds increased the preference
for less air
movement.](figs/supp_air_movement_acc_pref.png){#fig-supp-air-movement-acc-pref
width="75%"}

# References {.unnumbered}
